Equation-of-state insensitive relations after GW170817

### 2019-02-28 paper/bin-love-after-gw170817.tex
minor modification and final readthrough-spellcheck

The unconstrained set is then defined by drawing random samples in the spectral coefficients within their allowed prior ranges, and then eliminating any EoS that either leads to a polytropic index $\Gamma$ outside the allowed range, breaks the causality restriction, or breaks the maximum mass restriction. 
To obtain the constrained set of EoSs we analyze the publicly available data for GW170817 from the Gravitational Wave Open Science Center~\cite{GWOSC,Vallisneri:2014vxa}. We use the publicly available software library {\tt LALInference}~\cite{Veitch:2014wba,lalinference_o2} to obtain samples from the EoS posterior. Our anaysis uses the same settings and prior choice as those of~\cite{LIGO:posterior,Carney:2018sdv}. 
In both cases, each set consists of 100 members, with a subset of these shown in the left panel of Fig.~\ref{fig:eos}.

@@

\subsection{EoS-insensitive relations}\label{sec:eosInsensitive}
Current GW interferometry is not yet sensitive enough to accurately extract both tidal parameters $\tilde{\Lambda}$ and $\delta\tilde{\Lambda}$.
In a search to remedy this, Yagi and Yunes~\cite{Yagi:binLove} found that symmetric and antisymmetric combinations of the tidal deformabilities
\begin{equation}
\Lambda_s \equiv \frac{\Lambda_1 + \Lambda_2}{2}, \hspace{6mm} \Lambda_a \equiv \frac{\Lambda_1 - \Lambda_2}{2},
\end{equation}
display EoS-insensitive properties to a high degree, showing EoS variations of at most 20\% for binaries with masses less than $1.7 \text{ M}_{\odot}$ and using a representative sample of 11 EoSs. 
These ``binary Love relations" allow one to analytically break degeneracies between the tidal parameters: one can substitute $\Lambda_{a}=\Lambda_{a}(\Lambda_{s})$ in the GW model, thus completely eliminating $\Lambda_{a}$ from the parameter list (or vice-versa). This is important for two reasons: (i) the new model allows for the more accurate extraction of $\Lambda_{s}$ (or $\Lambda_{a}$ if $\Lambda_{s}$ is eliminated), and (ii) the relations allow for the inference of $\Lambda_{a}$ given a measurement of $\Lambda_{s}$ (or vice-versa), and from this for the inference of the individual tidal deformabilities $\Lambda_{1}$ and $\Lambda_{2}$. A simple Fisher analysis has shown that the binary Love relations improve parameter estimation of $\tilde{\Lambda}$ by up to an order of magnitude~\cite{Yagi:2015pkc,Yagi:binLove}.

@@

We can extract two additional conclusions from Fig.~\ref{fig:binLove}. First, as the mass ratio becomes more extreme (i.e.~away from unity), the ranges of values that $\Lambda_s$ and $\Lambda_a$ can take decrease. This is because as the separation between $m_1$ and $m_2$ increases, the effective number of possible $\Lambda_1(m_1)$-$\Lambda_2(m_2)$ configurations decrease, and thus, the allowed parameter space shrinks. Second, the fit to the constrained set does not extend to as high values of $\Lambda_s$ as the fit to the unconstrained set does. This is because of how the sets of EoSs were generated in our analysis, with the unconstrained set including some EoSs that are stiffer\footnote{Typically, the ``stiffness" of an EoS is determined by the amount of pressure gained given an increase in density. Stiff EoSs have steep pressures and predict more massive NSs, while soft EoSs have shallow pressures and predict smaller mass NSs.} and others that are softer than those constrained by the GW170817 event. This results in the unconstrained set  containing a larger range of tidal deformabilities $\Lambda_{1,2}$ (from both above and below) than the constrained set, which ultimately leads to a \emph{larger} range of $\Lambda_s=\frac{1}{2}(\Lambda_1+\Lambda_2)$ values.

@@

This behavior is exactly what we observe in Fig.~\ref{fig:binLove} for the binary Love relations of hybrid stars. Indeed, contrary to the case of EoS-insensitive relations for isolated hadronic stars (I-Love-Q, C-Love, and R-Love) which remain moderately EoS-insensitive for all hybrid stars, the binary Love relations depart from their hadronic counterparts below some transitional $\Lambda_{s}$. This is because at low pressures ($\Lambda_s>\Lambda_{s,\textrm{tr}}$) the binary is purely of NS/NS type (with the EoS for both stars a member of the constrained set), but once the critical pressure is reached ($\Lambda_{s}<\Lambda_{s,\textrm{tr}}$), one or both stars transition into the hybrid branch, resulting in large reductions in tidal deformability, as shown in the left panel of Fig.~\ref{fig:clove}. When one star lies on the hadronic-matter branch while the other is on the quark-matter branch, the difference between tidal deformabilities becomes larger than expected for pure hadronic-matter stars. This results in large deviations in the sums and differences of tidal deformabilities $(\Lambda_1 \pm \Lambda_2)$, disrupting the overall universality and generating the large ``bump" in the binary Love relations for hybrid stars\footnote{This discrepancy is not present in the I-Love-Q and C-Love relations due to their single-star nature.} seen in Fig~\ref{fig:binLove}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% APPENDIX B: MEASUREMENT ACCURACY AS A FUNCTION OF Q %%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\iffalse
\section{Measurement accuracy as a function of binary mass ratio}\label{app:measurement}
In this appendix, we discuss the effects of the binary NS mass ratio $q$ on the measurement accuracy of $\tilde\Lambda$ for various interferometers. 
The top panel of Fig.~\ref{fig:massVariation} displays an approximation of the $\tilde\Lambda$ measurement accuracy as a function of increasing mass ratio $q$ (for a fixed chirp mass of $\mathcal{M}=1.188 \text{ M}_{\odot}$ corresponding to GW170817), using the Fisher analysis techniques described in Sec.~\ref{sec:observations}.
We observe two points of interest about these trends: (i) the second generation detectors (O2, aLIGO, A\texttt{+}, and Voyager) typically become less accurate as the mass ratio approaches unity; and (ii) the third generation detectors (CE and ET) observe the opposite behavior-- a fast growth in accuracy as the mass ratio increases.
For the remainder of this appendix, we attempt to explain the opposing behaviors of the second and third generation interferometers on the extraction of $\tilde\Lambda$.

\begin{figure}
\begin{center} 
\includegraphics[width=.3\textwidth]{massVariation1.eps}
\includegraphics[width=.3\textwidth]{massVariation2.eps}
\includegraphics[width=.3\textwidth]{massVariation3.eps}
\end{center}
\caption{
Measurement accuracy of $\tilde\Lambda$ as a function of increasing mass ratio $q$ for interferometer sensitivities O2, aLIGO, A\texttt{+}, Voyager, CE, and ET.
Evaluated at a fixed chirp mass of $\mathcal{M}=1.188\text{ M}_{\odot}$ corresponding to GW170817, this is repeated for three cases: (i) (top) with correlations between all template waveform parameters $\theta^a$ intact; (ii) (center) with the correlations between $\tilde\Lambda$ and $\delta\tilde\Lambda$ removed; and (iii) (bottom) with correlations between all parameters removed.
The removal of parameter correlations is approximated by setting certain Fisher matrix elements to 0, as described in Appendix.~\ref{app:measurement}.
Observe how the top panel shows strong disagreement between second and third generation detectors, while the bottom panel shows identical behavior.
This indicates that parameter correlations between $\tilde\Lambda$ and higher PN order parameter $\delta\tilde\Lambda$ are the culprit in such a disagreement.
}
\label{fig:massVariation}
\end{figure}

We begin by considering the correlations between individual template parameters $\theta^a$ and $\tilde\Lambda$, given in Eq.~\ref{eq:template}.
A simple approximation for the amplitude of these correlations can be performed by assuming that all of the other correlations are non-existing.
Individual correlations between parameters $\tilde\Lambda$ and $\theta^a$ is achieved by setting all of the off-diagonal Fisher matrix elements to $0$, except for the $(i,j)$ components such that both $i$ and $j$ correspond to the $\tilde\Lambda$ and $\theta^a$ parameters.
We follow this process for each parameter in turn, and display the results in Fig.~\ref{fig:correlations} for each interferometer.
This figure, showing only the relative differences between correlations of different parameters $\theta^a$ and $\tilde\Lambda$, reveals a few things.
First, the coalescence time $t_c$ is highly correlated with $\tilde\Lambda$ equally for each interferometer, an unsurprising fact due to the high 4pN order at which the parameter enters the waveform.
Secondly, correlations with 6pN tidal parameter $\delta\tilde\Lambda$ shows high correlations with $\tilde\Lambda$ for only the third generation detectors.
This can be explained by the high pN order at which $\delta\tilde\Lambda$ enters the waveform, an effect that only the most sensitive of detectors can probe.

\begin{figure}
\begin{center} 
\begin{overpic}[width=\columnwidth]{correlations.eps}
\put(121,13){\fontsize{7pt}{7pt}\selectfont $\mathcal{M}$}
\end{overpic}
\end{center}
\caption{
Measurement accuracy of $\tilde\Lambda$ upon the consideration of correlations between \emph{only} $\tilde\Lambda$ and other individual template parameters $\theta^a$ shown in Eq.~\ref{eq:template}.
This process, repeated for 6 interferometer design sensitivities O2, aLIGO, A\texttt{+}, Voyager, CE, and ET, is approximated by setting all off-diagonal Fisher matrix elements to 0 except for the $(i,j)$ components such that both $i$ and $j$ correspond to the $\tilde\Lambda$ and $\theta^a$ parameters.
Observe how correlations with 4PN parameter $t_c$ (coalescence time) remain to be highly correlated with $\tilde\Lambda$ among all 6 interferometers.
However, correlations between 6PN tidal parameter $\delta\tilde\Lambda$ and $\tilde\Lambda$ only begin to make a strong impact for third generation detectors.
This is due to the small effect that $\delta\tilde\Lambda$ has on the waveform, only to be distinguished by the most sensitive detectors.
}
\label{fig:correlations}
\end{figure}

Following this conclusion on the difference of correlations between tidal parameters $\tilde\Lambda$ and $\delta\tilde\Lambda$ for second and third generation detectors, we attempt to describe the opposing behaviors of measurement accuracy between the two.
Shown in the central panel of Fig.~\ref{fig:massVariation}, we approximate the measurement accuracy on $\tilde\Lambda$ with the correlations between $\delta\tilde\Lambda$ and $\tilde\Lambda$ removed, through a process similar to that described above. 
Notice how the trends between detector generations become more similar -- implying that the correlations between tidal parameters may be the cause of the opposing behaviors.
This is confirmed by the bottom panel of Fig.~\ref{fig:massVariation} where all parameter correlations have been removed from the Fisher matrix.
We see that this results in identical behaviors between all 6 representative interferometers, leading us to conclude that the discrepancy between the interferometers originates from the degeneracies between different parameters in the gravitational waveform, namely the 6pN tidal parameter $\delta\tilde\Lambda$ which is only detectable by the highly sensitive third generation detectors CE and ET.
\fi
%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
%%%% BIBLIOGRAPHY %%%%
%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%

>>>>

The unconstrained set is then defined by drawing random samples in the spectral coefficients within their allowed prior ranges, and then eliminating any EoS that either leads to a polytropic index $\Gamma$ outside the allowed range, breaks the causality restriction, or breaks the maximum mass restriction. 
To obtain the constrained set of EoSs we analyze the publicly available data for GW170817 from the Gravitational Wave Open Science Center~\cite{GWOSC,Vallisneri:2014vxa}. We use the publicly available software library {\tt LALInference}~\cite{Veitch:2014wba,lalinference_o2} to obtain samples from the EoS posterior. Our analysis uses the same settings and prior choice as those of~\cite{LIGO:posterior,Carney:2018sdv}. 
In both cases, each set consists of 100 members, with a subset of these shown in the left panel of Fig.~\ref{fig:eos}.

@@

\subsection{EoS-insensitive relations}\label{sec:eosInsensitive}
Current GW interferometry is not yet sensitive enough to accurately extract both tidal parameters $\tilde{\Lambda}$ and $\delta\tilde{\Lambda}$.
In a search to remedy this, Yagi and Yunes~\cite{Yagi:binLove} found that symmetric and antisymmetric combinations of the tidal deformabilities
\begin{equation}
\Lambda_s \equiv \frac{\Lambda_2 + \Lambda_1}{2}, \hspace{6mm} \Lambda_a \equiv \frac{\Lambda_2 - \Lambda_1}{2},
\end{equation}
display EoS-insensitive properties to a high degree, showing EoS variations of at most 20\% for binaries with masses less than $1.7 \text{ M}_{\odot}$ and using a representative sample of 11 EoSs. 
These ``binary Love relations" allow one to analytically break degeneracies between the tidal parameters: one can substitute $\Lambda_{a}=\Lambda_{a}(\Lambda_{s})$ in the GW model, thus completely eliminating $\Lambda_{a}$ from the parameter list (or vice-versa). This is important for two reasons: (i) the new model allows for the more accurate extraction of $\Lambda_{s}$ (or $\Lambda_{a}$ if $\Lambda_{s}$ is eliminated), and (ii) the relations allow for the inference of $\Lambda_{a}$ given a measurement of $\Lambda_{s}$ (or vice-versa), and from this for the inference of the individual tidal deformabilities $\Lambda_{1}$ and $\Lambda_{2}$. A simple Fisher analysis has shown that the binary Love relations improve parameter estimation of $\tilde{\Lambda}$ by up to an order of magnitude~\cite{Yagi:2015pkc,Yagi:binLove}.

@@

We can extract two additional conclusions from Fig.~\ref{fig:binLove}. First, as the mass ratio becomes more extreme (i.e.~away from unity), the ranges of values that $\Lambda_s$ and $\Lambda_a$ can take decrease. This is because as the separation between $m_1$ and $m_2$ increases, the effective number of possible $\Lambda_1(m_1)$-$\Lambda_2(m_2)$ configurations decrease, and thus, the allowed parameter space shrinks. Second, the fit to the constrained set does not extend to as high values of $\Lambda_s$ as the fit to the unconstrained set does. This is because of how the sets of EoSs were generated in our analysis, with the unconstrained set including some EoSs that are stiffer\footnote{Typically, the ``stiffness" of an EoS is determined by the amount of pressure gained given an increase in density. Stiff EoSs have steep pressures and predict larger maximum mass NSs, while soft EoSs have shallow pressures and predict smaller maximum mass NSs.} and others that are softer than those constrained by the GW170817 event. This results in the unconstrained set  containing a larger range of tidal deformabilities $\Lambda_{1,2}$ (from both above and below) than the constrained set, which ultimately leads to a \emph{larger} range of $\Lambda_s=\frac{1}{2}(\Lambda_1+\Lambda_2)$ values.

@@

This behavior is exactly what we observe in Fig.~\ref{fig:binLove} for the binary Love relations of hybrid stars. Indeed, contrary to the case of EoS-insensitive relations for isolated hadronic stars (I-Love-Q, C-Love, and R-Love) which remain moderately EoS-insensitive for all hybrid stars, the binary Love relations depart from their hadronic counterparts below some transitional $\Lambda_{s}$. This is because at low pressures ($\Lambda_s>\Lambda_{s,\textrm{tr}}$) the binary is purely of NS/NS type (with the EoS for both stars a member of the constrained set), but once the critical pressure is reached ($\Lambda_{s}<\Lambda_{s,\textrm{tr}}$), one or both stars transition into the hybrid branch, resulting in large reductions in tidal deformability, as shown in the left panel of Fig.~\ref{fig:clove}. When one star lies on the hadronic-matter branch while the other is on the quark-matter branch, the difference between tidal deformabilities becomes larger than expected for pure hadronic-matter stars. This results in large deviations in the sums and differences of tidal deformabilities $(\Lambda_2 \pm \Lambda_1)$, disrupting the overall universality and generating the large ``bump" in the binary Love relations for hybrid stars\footnote{This discrepancy is not present in the I-Love-Q and C-Love relations due to their single-star nature.} seen in Fig~\ref{fig:binLove}.

@@

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
%%%% BIBLIOGRAPHY %%%%
%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
